Multi-reference pangenome graphs

### 2022-01-05 manuscript/main.tex
onward

\section{Model}
\label{sec:model}

@@

A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Garrison_2019_thesis,Eizenga_2020}.
In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
Each node $n_i$ has a unique numerical identifier $i$ and an implicit reverse complement $\bar{n_i}$.
A node strand $s$ corresponds to one node orientation.
Edges $E = e_1\ldots e_{|E|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written terms of the sequence graph.

A variation graph models a set of pairwise alignments between its embedded paths.
The variation graph shows that two sequences match when they both ``step'' on the same node or series of nodes.
This is trivial given our definition.
If two sub-paths contain the same set of steps, they must have identical subsequences.
%Thus, to represent alignments in the graph, we built it by rendering matches into the graph
We can construct a variation graph from a set of pairwise alignments by guaranteeing that each matching pair of subsequences are represented in a single walk in the output graph.

@@

\subsection{Sequence index}

As input, we have $Q$, which is a concatenation of the sequences from which we will build the graph.
We build a compressed suffix array (CSA) mapping sequence names to offsets in $Q$, and also the inverse using a rank/select dictionary on a bitvector marking the starts of sequences in $Q$.
This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.

@@

>>>>

\section{Variation graph induction}
\label{sec:model}

@@

A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Garrison_2019_thesis,Eizenga_2020}.
In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
Each node $n_i$ has a unique numerical identifier $i$ and an implicit reverse complement $\bar{n_i}$.
A node strand $s$ corresponds to one node orientation.
Edges $E = e_1\ldots e_{|E|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written in terms of the sequence graph.

A variation graph represents pairwise alignments between its embedded paths.
We infer that two sequences match when they both ``step'' on the same node or series of nodes.
This follows from our definition of the graph itself.
If two sub-paths contain the same set of steps, they must have identical subsequences.
%Thus, to represent alignments in the graph, we built it by rendering matches into the graph
%The core of our graph constr TODO HERE
We can construct a variation graph from a set of pairwise alignments by guaranteeing that each matching pair of subsequences are represented in a single walk in the output graph.

@@

\subsection{Sequence input}

As input, we have $Q = g_1 \ldots g_N$, which is a concatenation of the $N$ genome sequences from which we will build the graph.
Using a compressed suffix array (CSA) over sequence names, we build an index that maps between offsets in all genomes and offsets in $Q$ via lookup of the sequence name in the CSA.
This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.

@@

Alignments between genomes imply a set of exact matches between subsequences of $Q$, which we represent as interval pairs $m := ((q_b, q_e), (t_b, t_e))$.
We encode the $X$ matches in a set of alignments made by an external method \cite{wfmash,Li_2018} in $A = m_1 \ldots m_X$.

\citep{Li_bedtk_2020}.
